Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-04-22 co-regular-EPQ.tex
some stuff

Section~\ref{ssec:aut} describes a $\mathsf{MatchedMaxPriority}$-complete automaton, the other automata being defined in Appendix~\ref{sec:appendix lemma and register automata for FIFO of single-priority executions}, %Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}, 
%Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Equal} 
and Appendix \ref{subsec:co-regular of EPQ3}. Therefore, the following holds.

@@

Also, we consider $\Gamma$-complete automata for $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$, recognizing executions which contain only one maximal priority. This is possible because any data-differentiated execution for which $\Gamma(e)$ is false has such a projection.
%Also, we consider $\Gamma$-complete automata for $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$, recognizing executions where any two occurring priorities are comparable (w.r.t. $\prec$). This is possible because any data-differentiated execution which is not $\Gamma$-linearizable has such a projection.
Formally, given a data-differentiated execution $e$ and $p$ a maximal priority in $e$, $e\vert_{\preceq p}$ is the projection of $e$ to the set of values with priorities comparable to $p$. Then,

@@

The following result shows that $\Gamma$-complete automata enable an effective reduction of checking linearizability of concurrent priority queue implementations to state reachability. It is a direct consequence of the above definitions.

\begin{restatable}{theorem}{ReduceEPQIntoStateReachability}
\label{lemma:reduce EPQ into state reachability}
Let $\mathcal{I}$ be a data-independent implementation, and $A(\Gamma)$ a $\Gamma$-complete automaton for each $\Gamma$. Then,
$\mathcal{I} \sqsubseteq \seqPQ$ if and only if $\mathcal{I} \cap A(\Gamma) = \emptyset$ for all $\Gamma$.
\end{restatable}

Section~\ref{subsec:combine step-by-step linearizability and co-regular} discusses decidability results implied by the theorem above.

@@

A differentiated execution $e$ is not $\mathsf{MatchedMaxPriority}$-linearizable when all the $\textit{put}$ operations in $e$ using the maximal priority $p$ are matched, and $e$ is not linearizable w.r.t. the set of sequential executions satisfying $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$ for each value $x$ of priority $p$. We consider two cases depending on whether $e$ contains exactly one value with priority $p$ or at least two values. We denote by $\mathsf{MatchedMaxPriority}^{>}$ the strengthening of $\mathsf{MatchedMaxPriority}$ with the condition that all the values other than $x$ have a priority strictly smaller than $p$ (corresponding to the first case), and by $\mathsf{MatchedMaxPriority}^{=}$ the strengthening of the same formula with the negation of this condition (corresponding to the second case).
We use particular instances of register automata~\cite{finite-memory, the paper I found} whose states include only two registers, one for storing a priority guessed at the initial state, and one for storing the priority of the current action in the execution. The transitions can check equality or the order relation $\prec$ between the values stored in the two registers. Instead of formalizing the full class of register automata, we consider a simpler class which suffices our needs. More precisely, we consider a class of labeled transition systems whose states consist of a finite control part and a register $r$ interpreted to elements of $\mathbb{P}$. The transition labels can be one of the following:
\begin{itemize}
	\item $r=*$ for storing an arbitrary value to $r$,
	\item $\textit{call}(\textit{rm},a)$ and $\textit{ret}(\textit{rm},a)$ for reading call/return actions of a remove,
	\item $\textit{call}(\textit{put},d,g)$ where $g\in\{=r,\prec r,true\}$ is a guard, for reading a call action $\textit{call}(\textit{put},d,p)$ of a put and checking whether $p$ is either equal to or smaller than the value stored in $r$, or arbitrary,
	\item $\textit{ret}(\textit{put},d,true)$ for reading a return action $\textit{ret}(\textit{put},d,p)$ for some $p$.
\end{itemize}
The set of words accepted by such a transition system can be defined as usual.

@@

The execution in \figurename~\ref{fig:introduce gap for EPQ1Lar} is not $\mathsf{MatchedMaxPriority}^>$-linearizable because the left-right constraint of the maximal priority value $b$ contains a cycle: $f \rightarrow e \rightarrow d \rightarrow b \rightarrow f$. We prove that the presence of such a cycle is equivalent to the execution not being $\mathsf{MatchedMaxPriority}^>$-linearizable (see Appendix~\ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}).
%The existence of cycle though value with maximal priority in left-right constraint formally modelled the phenomenon in \figurename~\ref{fig:introduce gap for EPQ1Lar}. For example, in \figurename~\ref{fig:introduce gap for EPQ1Lar}, $f \rightarrow e \rightarrow d \rightarrow b \rightarrow f$ is a cycle in left-right constraint. We then prove that getting rid of cycle though value with maximal priority in left-right constraint is enough for ensure linearizable w.r.t $\mathsf{MatchedMaxPriority}^{>}$.
When the left-right constraint of the maximal priority value $x$ contains a cycle of the form $d_1 \rightarrow \ldots \rightarrow d_m \rightarrow x \rightarrow d_1$ for some $d_1$,$\ldots$,$d_n\in \mathbb{D}$, we say that $x$ is \emph{covered} by $d_1,\ldots,d_m$. The shape of such a cycle (i.e., the alternation between call/return actions of $\textit{put}$/$\textit{rm}$ operations) can be detected using our class of automata, the only complication being the unbounded number of values $d_1$,$\ldots$,$d_n$. However, by data independence, whenever an implementation contains such an execution it also contains an execution where all the values $d_1$,$\ldots$,$d_n$ are renamed to the same value $a$, and $x$ is renamed to $b$. Therefore, our automata can be defined over a fixed set of values $a$, $b$, and $\top$ (for the operations outside of the non-linearizable projection).

%TODO WHY IT IS A SET OF AUTOMATA ? SAY WHAT ARE THE CASES THAT NEED TO BE COVERED. SAY WHAT IS THE CASE WE GIVE IN THE FIGURE. USE $\top$ AS MENTIONED AT THE BEGINNING OF THE SECTION INSTEAD OF 	$c$. USE BIG CAPS FOR SETS OF TRANSITION LABELS.
To define a $\mathsf{MatchedMaxPriority}^>$-complete automaton we need to consider several cases depending on the order between the call/return actions of the $\textit{put}$/$\textit{rm}$ operations that add and respectively, remove the value $b$. For example, the case where the put happens-before the remove (as in \figurename~\ref{fig:introduce gap for EPQ1Lar}) is pictured in \figurename~\ref{fig:automata APQ1Lar-1 in paper}. This automaton captures the three possible ways of ordering the first action $\textit{ret}(\textit{put},a,\_)$ w.r.t. the actions with value $b$, which are pictured in \figurename~\ref{fig:executions APQ1Lar-1 in paper} (this action cannot occur after $\textit{call}(\textit{rm},b,\_)$ since $b$ must be covered by the $a$s). The paths corresponding to these three possible orders are: $q_1 \rightarrow q_2 \rightarrow q_3 \ldots \rightarrow q_7$, $q_1 \rightarrow q_2 \rightarrow q_3 \ldots \rightarrow q_{10}$ and $q_1 \rightarrow q_9 \rightarrow q_{10} \ldots \rightarrow q_7$. Automata for the other cases can be found in Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}.

@@

The case of \figurename~\ref{fig:introduce pb order} can be formally described as follows: $a <_{\textit{pb}}^* b$ while the right-most gap-point of $b$ is before $\textit{call}(\textit{rm},a)$ or $\textit{call}(\textit{put},a,p_4)$. The following lemma states that these conditions are enough to characterize non-linearizability w.r.t $\mathsf{MatchedMaxPriority}^{=}$.

@@

\begin {proof} (Sketch)
We have already intuitively explain the proof of the $\textit{if}$ direction. To prove the $\textit{only if}$ direction, we prove its contrapositive. %, where we already know that $\forall x, y$ with maximal priority, if $y <_{\textit{pb}}^* x$, then the rightmost gap-point of $x$ is after $\textit{cal}(\textit{put},y,\textit{pri})$ and $\textit{cal}(\textit{rm},x)$. We need to prove that $e \sqsubseteq \textit{MS}(\textit{EPQ}_1^{=})$. %Or we can say, we need to explicitly construct linearization of $e$.

Let $\textit{Items}(e,p)$ be the set of values with priority $p$ in $e$. We introduce another lemma (Lemma \ref{lemma:maximal in pb and gap-point make a candidate of EPQ1Equal} in Appendix), which states that if we can find a value $\exists g_1 \in \textit{Items}(e,p)$, such that $\forall g_2 \in \textit{Items}(e,p)$, (1) $g_1 \nless_{\textit{pb}} g_2$, and (2) the right-most gap-point of $g_1$ is after $\textit{call}(\textit{put},g_2,p)$ and $\textit{call}(\textit{rm},g_2)$. Then $e$ is linearizable w.r.t $\mathsf{MatchedMaxPriority}^{=}$.
The remaining of our proof is a loop for searching such $g_1$ in $e$. Let $e_p$ be the projection of $e$ into values of priority $p$, and let $l_p$ be the linearization of $e$ which satisfy the requirements of queue. Let the initial value of $a_1$ be the last inserted value in $l_p$.

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] check if $a_1$ satisfy the conditions of Lemma \ref{lemma:maximal in pb and gap-point make a candidate of EPQ1Equal}. If it does, then this loop terminates.

\item[-] Otherwise, $\exists a_2 \in \textit{Items}(e,p)$, such that the rightmost gap-point of $a_1$ is before $\textit{cal}(\textit{put},a_2,p)$ or $\textit{cal}(\textit{rm},a_2)$ in $e$. By assumption, we know that $a_2 \nless_{\textit{pb}} a_1$, and each gap-point of $a_2$ is after the rightmost gap-point of $a_1$.

    \begin{itemize}
    \setlength{\itemsep}{0.5pt}
    \item[-] If $\forall b \in \textit{Items}(e,p)$, $a_2\not<_{\textit{pb}}b$. Then we go to next round of the loop and make $a_2$ to be the ``new-round $a_1$''.
    \item[-] Otherwise, there exists $a_3 \in \textit{Items}(e,p)$ such that $a_2 <_{\textit{pb}}^* a_3$. It is not hard to see that there is no cycle in $<_{\textit{pb}}$, and it is safe to assume that $a_3$ is maximal w.r.t $<_{\textit{pb}}^*$.

        By assumption,we know that the rightmost gap-point of $a_3$ is after $\textit{call}(\textit{put},a_2,p)$ and $\textit{call}(\textit{rm},a_2)$. Therefore, we can see that the rightmost gap-point of $a_3$ is after the rightmost gap-point of $a_1$. Then we go to next round of the loop and make $a_3$ to be the ``new-round $a_1$''.
    \end{itemize}
\end{itemize}

Let $a^i$ be the $a_1$ in the $\textit{i-th}$ round of loop. We can see that the rightmost gap-point of $a^j$ is after the rightmost gap-point of $a^i$ for each $i<j$. Therefore, the loop finally terminates at some $a^f$ that satisfies the condition of Lemma \ref{lemma:maximal in pb and gap-point make a candidate of EPQ1Equal}. This implies that $e$ is linearizable w.r.t $\mathsf{MatchedMaxPriority}^{=}$ and completes the proof of the $\textit{only if}$ direction.
\end {proof}

>>>>

Section~\ref{ssec:aut} describes a $\mathsf{MatchedMaxPriority}$-complete automaton, the other automata being defined in Appendix~\ref{sec:appendix lemma and register automata for FIFO of single-priority executions} %Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}, 
%Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Equal} 
and Appendix \ref{subsec:co-regular of EPQ3}. Therefore, the following holds.

@@

Also, we consider $\Gamma$-complete automata for $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$, recognizing executions which contain only one maximal priority. This is possible because any data-differentiated execution for which $\Gamma(e)$ is false has such a projection.
%Also, we consider $\Gamma$-complete automata for $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$, recognizing executions where any two occurring priorities are comparable (w.r.t. $\prec$). This is possible because any data-differentiated execution which is not $\Gamma$-linearizable has such a projection.
Formally, given a data-differentiated execution $e$ and $p$ a maximal priority in $e$, $e\vert_{\preceq p}$ is the projection of $e$ to the set of values with priorities smaller than $p$. Then,

@@

The following shows that $\Gamma$-complete automata enable an effective reduction of checking linearizability of concurrent priority queue implementations to state reachability. It is a direct consequence of the above definitions. Section~\ref{subsec:combine step-by-step linearizability and co-regular} discusses decidability results implied by this reduction.

\begin{restatable}{theorem}{ReduceEPQIntoStateReachability}
\label{lemma:reduce EPQ into state reachability}
Let $\mathcal{I}$ be a data-independent implementation, and $A(\Gamma)$ be a $\Gamma$-complete automaton for each $\Gamma$. Then,
$\mathcal{I} \sqsubseteq \seqPQ$ if and only if $\mathcal{I} \cap A(\Gamma) = \emptyset$ for all $\Gamma$.
\end{restatable}

@@

A differentiated execution $e$ is not $\mathsf{MatchedMaxPriority}$-linearizable when all the $\textit{put}$ operations in $e$ using the maximal priority $p$ are matched, and $e$ is not linearizable w.r.t. the set of sequential executions satisfying $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$ for each value $x$ of priority $p$. We consider two cases depending on whether $e$ contains exactly one value with priority $p$ or at least two values. We denote by $\mathsf{MatchedMaxPriority}^{>}$ the strengthening of $\mathsf{MatchedMaxPriority}$ with the condition that all the values other than $x$ have a priority strictly smaller than $p$ (corresponding to the first case), and by $\mathsf{MatchedMaxPriority}^{=}$ the strengthening of the same formula with the negation of this condition (corresponding to the second case).
We use particular instances of register automata~\cite{DBLP:journals/tcs/KaminskiF94,DBLP:conf/icalp/Cerans94,DBLP:conf/stacs/SegoufinT11} whose states include only two registers, one for storing a priority guessed at the initial state, and one for storing the priority of the current action in the execution. The transitions can check equality or the order relation $\prec$ between the values stored in the two registers. Instead of formalizing the full class of register automata, we consider a simpler class which suffices our needs. More precisely, we consider a class of labeled transition systems whose states consist of a finite control part and a register $r$ interpreted to elements of $\mathbb{P}$. The transition labels can be one of the following:
\begin{itemize}
	\item $r=*$ for storing an arbitrary value to $r$,
	\item $\textit{call}(\textit{rm},a)$ and $\textit{ret}(\textit{rm},a)$ for reading call/return actions of a remove,
	\item $\textit{call}(\textit{put},d,g)$ where $g\in\{=r,\prec r,true\}$ is a guard, for reading a call action $\textit{call}(\textit{put},d,p)$ of a put and checking whether $p$ is either equal to or smaller than the value stored in $r$, or arbitrary,
	\item $\textit{ret}(\textit{put},d,true)$ for reading a return action $\textit{ret}(\textit{put},d,p)$ for any $p$.
\end{itemize}
The set of words accepted by such a transition system can be defined as usual.

@@

The execution in \figurename~\ref{fig:introduce gap for EPQ1Lar} is not $\mathsf{MatchedMaxPriority}^>$-linearizable because the left-right constraint of the maximal priority value $b$ contains a cycle: $f \rightarrow e \rightarrow d \rightarrow b \rightarrow f$. The presence of such a cycle is equivalent to the execution not being $\mathsf{MatchedMaxPriority}^>$-linearizable (see Appendix~\ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}).
%The existence of cycle though value with maximal priority in left-right constraint formally modelled the phenomenon in \figurename~\ref{fig:introduce gap for EPQ1Lar}. For example, in \figurename~\ref{fig:introduce gap for EPQ1Lar}, $f \rightarrow e \rightarrow d \rightarrow b \rightarrow f$ is a cycle in left-right constraint. We then prove that getting rid of cycle though value with maximal priority in left-right constraint is enough for ensure linearizable w.r.t $\mathsf{MatchedMaxPriority}^{>}$.
When the left-right constraint of the maximal priority value $x$ contains a cycle of the form $d_1 \rightarrow \ldots \rightarrow d_m \rightarrow x \rightarrow d_1$ for some $d_1$,$\ldots$,$d_n\in \mathbb{D}$, we say that $x$ is \emph{covered} by $d_1,\ldots,d_m$. The shape of such a cycle (i.e., the alternation between call/return actions of $\textit{put}$/$\textit{rm}$ operations) can be detected using our class of automata, the only complication being the unbounded number of values $d_1$,$\ldots$,$d_n$. However, by data independence, whenever an implementation contains such an execution it also contains an execution where all the values $d_1$,$\ldots$,$d_n$ are renamed to the same value $a$, and $x$ is renamed to $b$. Therefore, our automata can be defined over a fixed set of values $a$, $b$, and $\top$ (recall that $\top$ is used for the operations outside of the non-linearizable projection).

%TODO WHY IT IS A SET OF AUTOMATA ? SAY WHAT ARE THE CASES THAT NEED TO BE COVERED. SAY WHAT IS THE CASE WE GIVE IN THE FIGURE. USE $\top$ AS MENTIONED AT THE BEGINNING OF THE SECTION INSTEAD OF 	$c$. USE BIG CAPS FOR SETS OF TRANSITION LABELS.
To define a $\mathsf{MatchedMaxPriority}^>$-complete automaton we need to consider several cases depending on the order between the call/return actions of the $\textit{put}$/$\textit{rm}$ operations that add and respectively, remove the value $b$. For example, the case where the put happens-before the remove (as in \figurename~\ref{fig:introduce gap for EPQ1Lar}) is pictured in \figurename~\ref{fig:automata APQ1Lar-1 in paper}. This automaton captures the three possible ways of ordering the first action $\textit{ret}(\textit{put},a,\_)$ w.r.t. the actions with value $b$, which are pictured in \figurename~\ref{fig:executions APQ1Lar-1 in paper} (this action cannot occur after $\textit{call}(\textit{rm},b,\_)$ since $b$ must be covered by the $a$-s). The paths corresponding to these three possible orders are: $q_1 \rightarrow q_2 \rightarrow q_3 \ldots \rightarrow q_7$, $q_1 \rightarrow q_2 \rightarrow q_3 \ldots \rightarrow q_{10}$ and $q_1 \rightarrow q_9 \rightarrow q_{10} \ldots \rightarrow q_7$. Automata for the other cases can be found in Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}.

@@

The case of \figurename~\ref{fig:introduce pb order} can be formally described as follows: $a <_{\textit{pb}}^* b$ while the right-most gap-point of $b$ is before $\textit{call}(\textit{rm},a)$ or $\textit{call}(\textit{put},a,p_4)$. The following lemma states that these conditions are enough to characterize non-linearizability w.r.t $\mathsf{MatchedMaxPriority}^{=}$ (see Appendix\ref{sec:appendix proof and definition in section co-regular of EPQ1Equal}).

@@

%\begin {proof} (Sketch)
%We have already gave an intuitive explanation for the $\textit{if}$ direction. To prove the $\textit{only if}$ direction, we prove its contrapositive. %, where we already know that $\forall x, y$ with maximal priority, if $y <_{\textit{pb}}^* x$, then the rightmost gap-point of $x$ is after $\textit{cal}(\textit{put},y,\textit{pri})$ and $\textit{cal}(\textit{rm},x)$. We need to prove that $e \sqsubseteq \textit{MS}(\textit{EPQ}_1^{=})$. %Or we can say, we need to explicitly construct linearization of $e$.
%
%Let $\textit{values}(e,p)$ be the set of values with priority $p$ in $e$. We introduce another lemma (Lemma \ref{lemma:maximal in pb and gap-point make a candidate of EPQ1Equal} in Appendix), which states that if we can find a value $\exists g_1 \in \textit{values}(e,p)$, such that $\forall g_2 \in \textit{values}(e,p)$, (1) $g_1 \nless_{\textit{pb}} g_2$, and (2) the right-most gap-point of $g_1$ is after $\textit{call}(\textit{put},g_2,p)$ and $\textit{call}(\textit{rm},g_2)$. Then $e$ is linearizable w.r.t $\mathsf{MatchedMaxPriority}^{=}$.
%The remaining of our proof is a loop for searching such $g_1$ in $e$. Let $e_p$ be the projection of $e$ into values of priority $p$, and let $l_p$ be the linearization of $e$ which satisfy the requirements of queue. Let the initial value of $a_1$ be the last inserted value in $l_p$.
%
%\begin{itemize}
%\setlength{\itemsep}{0.5pt}
%\item[-] check if $a_1$ satisfy the conditions of Lemma \ref{lemma:maximal in pb and gap-point make a candidate of EPQ1Equal}. If it does, then this loop terminates.
%
%\item[-] Otherwise, $\exists a_2 \in \textit{values}(e,p)$, such that the rightmost gap-point of $a_1$ is before $\textit{cal}(\textit{put},a_2,p)$ or $\textit{cal}(\textit{rm},a_2)$ in $e$. By assumption, we know that $a_2 \nless_{\textit{pb}} a_1$, and each gap-point of $a_2$ is after the rightmost gap-point of $a_1$.
%
%    \begin{itemize}
%    \setlength{\itemsep}{0.5pt}
%    \item[-] If $\forall b \in \textit{values}(e,p)$, $a_2\not<_{\textit{pb}}b$. Then we go to next round of the loop and make $a_2$ to be the ``new-round $a_1$''.
%    \item[-] Otherwise, there exists $a_3 \in \textit{values}(e,p)$ such that $a_2 <_{\textit{pb}}^* a_3$. It is not hard to see that there is no cycle in $<_{\textit{pb}}$, and it is safe to assume that $a_3$ is maximal w.r.t $<_{\textit{pb}}^*$.
%
%        By assumption,we know that the rightmost gap-point of $a_3$ is after $\textit{call}(\textit{put},a_2,p)$ and $\textit{call}(\textit{rm},a_2)$. Therefore, we can see that the rightmost gap-point of $a_3$ is after the rightmost gap-point of $a_1$. Then we go to next round of the loop and make $a_3$ to be the ``new-round $a_1$''.
%    \end{itemize}
%\end{itemize}
%
%Let $a^i$ be the $a_1$ in the $\textit{i-th}$ round of loop. We can see that the rightmost gap-point of $a^j$ is after the rightmost gap-point of $a^i$ for each $i<j$. Therefore, the loop finally terminates at some $a^f$ that satisfies the condition of Lemma \ref{lemma:maximal in pb and gap-point make a candidate of EPQ1Equal}. This implies that $e$ is linearizable w.r.t $\mathsf{MatchedMaxPriority}^{=}$ and completes the proof of the $\textit{only if}$ direction.
%\end {proof}
